---
title: "Cetaceans Species Richness Analysis"
description: |
  Mapping species richness of California cetacean species.
author:
  - name: "Roshni Katrak-Adefowora"
    url: {}
date: 02-25-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(tidyverse)
library(here)
library(sf)
library(rnaturalearth)
```

Initial Data Wrangling

```{r, results = FALSE}
#read in the data
cetaceans <- list.files(path = here("_posts", "2021-02-25-cetaceans", "ca_cetaceans"), full.names = TRUE)
cetaceans_raster <- raster::stack(cetaceans)

#get coastline for CA
ne_ca <- ne_download(scale = "medium", type = "states", category = "cultural", returnclass = "sf") %>% 
  filter(name == "California")

#make threshold function
threshold_present <- function(x) {
  y <- ifelse(x >= 0.6, 1, 0) #if >= 0.6, then 1, if not then 0
  return(y)
}

#use threshold function
species_threshold <- calc(cetaceans_raster, fun = threshold_present)

#find number of species in each cell
species_richness <- calc(species_threshold, fun = sum, na.rm = TRUE)

#convert raster to data frame for ggplot
species_richness_df <- raster::rasterToPoints(species_richness) %>% #convert to matrix
  as.data.frame() %>% #convert to dataframe
  filter(layer > 0) #remove zeros
```

Plot Species Richness

```{r}
ggplot()+
  geom_raster(data = species_richness_df, aes(x = x, y = y, fill = layer))+
  geom_sf(data = ne_ca)+
  coord_sf(xlim = c(-125,-115), ylim = c(32,38))+ #set x and y limits of graph
  scale_fill_gradient(low = 'violetred', high = 'purple4')+
  theme_light()+
  labs(x="Longitude",
       y="Latitude",
       fill="Species Richness")
```

**Figure 1. Cetacean species richness in the California Bight.** For 35 cetacean species, the probability of occurence was used to find species richness. The presence of each species was determined using a threshold value 0f 0.6. Data: AquaMaps (2016).

**Citation:** AquaMaps (Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016). AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org

